---
title: "Land Cover Change Detection in Southwest Minnesota"
author: Olivia Adomabea
subtitle: Subtitle here if desired
date: 11/22/2024
date-format: long
---

# Introduction
This project seeks to detect land cover changes in southern Minnesota. Understanding the dynamics of environmental transformations caused by natural processes and human activities requires detecting land cover change. Accurate detection helps inform sustainable land-use policies and mitigate environmental impacts, such as biodiversity decline and climate change. Land cover change detection is a critical tool for understanding the complex dynamics of land use and its impacts on the environment, society, and economy. As human activities, such as urban expansion, agricultural intensification, and deforestation, continue to alter the landscape, it becomes increasingly important to monitor and assess these changes over time. studying land cover changes is vital for understanding environmental, social, and economic systems. It informs policies, promotes sustainable development, enhances disaster resilience, and helps manage natural resources more effectively.This study attempt to answer the principal questions:
-What are the land cover change patterns that have occurred in Southern Minnesota from 2000 to 2020?
-What factors are causing these chanhes if there has been any change detection?

# Materials and methods
The data source and data processing and analyzing method are noted in the code chunk in r. 
Packages installed/loaded:
-tidyverse
-terra
-RColorBrewer
Data Source:
-Landsat 8 (path 28. row 29)
-copied path of downloaded data.

Code: The code associated with the project is well organized and easy to follow. Demonstrates mastery of R graphics and functions.

Data: The underlying data are publicly accessible via the web and downloaded/accessed within the Rmd script. 
---

```{r}
# load packages
# install.packages("terra")
library(terra)
library(tidyverse)
library(RColorBrewer)
```

```{r}
# load raster file
landcover_path_2000 <- "/Users/oliviaadomabea/Documents/Spatial_Data_Analysis_Respo/case-studies-Oliviaado/Final Project/data/MNLC/landcover_2000.tif"

landcover_path_2020 <- "/Users/oliviaadomabea/Documents/Spatial_Data_Analysis_Respo/case-studies-Oliviaado/Final Project/data/MNLC/landcover_2020.tif"

landcover_2000 <- rast(landcover_path_2000)
landcover_2020 <- rast(landcover_path_2020)
print(landcover_2000)
print(landcover_2020)
```

```{r}
# explore raster file
landcover_cat_2000 <- unique(values(landcover_2000))
landcover_cat_2020 <- unique(values(landcover_2020))
print(landcover_cat_2000)
print(landcover_cat_2020)
```

```{r}
# count pixels per landcover category
landcover_2000_pc <- freq(landcover_2000, digits = 0)
print(landcover_2000_pc)
landcover_2020_pc <- freq(landcover_2020, digits = 0)
print(landcover_2020_pc)
```

```{r}
# for 2000
# convert pixel counts to a df
landcover_2000_pc <- as.data.frame(landcover_2000_pc[, c(2,3)])
colnames(landcover_2000_pc) <- c("Landcover_Class", "Pixel_Count_2000")
print(landcover_2000_pc)

# adding proportion to the df
total_pixels <- sum(landcover_2000_pc$Pixel_Count_2000)
landcover_2000_pc$Percentage_2000 <- (landcover_2000_pc$Pixel_Count_2000 / total_pixels) * 100
```

```{r}
# for 2020
# convert pixel counts to a df
landcover_2020_pc <- as.data.frame(landcover_2020_pc[, c(2,3)])
colnames(landcover_2020_pc) <- c("Landcover_Class", "Pixel_Count_2020")
print(landcover_2020_pc)

# adding proportion to the df
total_pixels <- sum(landcover_2020_pc$Pixel_Count_2020)
landcover_2020_pc$Percentage_2020 <- (landcover_2020_pc$Pixel_Count_2020 / total_pixels) * 100
```

```{r}
library(tidyverse)
lc_2000_2020 <- left_join(landcover_2000_pc, landcover_2020_pc, by = "Landcover_Class")
lc_2000_2020 <- lc_2000_2020 %>%
  mutate(land_cover_class = case_when(
    Landcover_Class == 11 ~ "open water",
    Landcover_Class == 21 ~ "developed area",
    Landcover_Class == 22 ~ "developed area",
    Landcover_Class == 23 ~ "developed area",
    Landcover_Class == 24 ~ "developed area",
    Landcover_Class == 31 ~ "barren land",
    Landcover_Class == 41 ~ "forest",
    Landcover_Class == 42 ~ "forest",
    Landcover_Class == 43 ~ "forest",
    Landcover_Class == 52 ~ "grassland",
    Landcover_Class == 71 ~ "grassland",
    Landcover_Class == 81 ~ "grassland",
    Landcover_Class == 82 ~ "cultivated crops",
    Landcover_Class == 90 ~ "wetlands",
    Landcover_Class == 95 ~ "wetlands"
  ))

lc_2000_2020 <- lc_2000_2020[, c(2:6)]
lc_2000_2020 <- lc_2000_2020 %>% group_by(land_cover_class) %>%
  summarise(across(everything(), sum, na.rm = TRUE))
```

```{r}
# some exploratory data analysis
# 1. group bar plot
data <- lc_2000_2020 %>% 
  pivot_longer(cols = c(Percentage_2000, Percentage_2020), 
               names_to = "Year", values_to = "Value")

ggplot(data, aes(x = land_cover_class, y = Value, fill = Year)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(title = "Landcover Comparison by Year",
       x = "Land Cover",
       y = "Percent Coverage") +
  theme_minimal() + scale_fill_brewer(palette = "Set3")
```

```{r}
# 2. visualizing the raster file spatially
n_classes <- length(landcover_2000_pc$Landcover_Class)
palette <- brewer.pal(n = n_classes, name = "Set3")

plot(landcover_2000, col = palette, legend = TRUE,
     main = "Spatial Distribution of Landcover Categories\n of Southwestern Minnesota in 2000")
```
```{r}
plot(landcover_2020, col = palette, legend = TRUE,
     main = "Spatial Distribution of Landcover Categories\n of Southwestern Minnesota in 2020")
```

```{r}
# preprocessing for change detection
landcover_2020 <- resample(landcover_2020, landcover_2000, method = "near")

# calculating change
change <- landcover_2020 - landcover_2000
```

```{r}
change[change == 0] <- NA
custom_colors <- c("white", "green", "blue", "yellow", "red", "purple")

# visualize change
plot(change, col = custom_colors[-1], legend = TRUE,
     main = "Landcover Changes (2000 - 2020)")
```
# Conclusions

\[\~200 words\]

Clear summary adequately describing the results and putting them in context. Discussion of further questions and ways to continue investigation.

# References

All sources are cited in a consistent manner
